### 2020-07-03 manuscript/manuscript.rmd
second pass through text

There has been considerable recent debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenomic analyses find support for a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others found support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in a many respects, including the genes and species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 13 previous phylogenomic analyses and performing new standardized analyses. Our reanalyses are consistent with previous studies that recover Porifera-sister with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative. Since then many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1). These studies have taken a wide variety of approaches and results have varied across these different analyses. The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result. But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses. This methodological variance is often confounded, for example when a particular model is only available in one software package.

To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Extended Data Table 1). These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.

![](figures/Figure_overview.png)
**Fig. 1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta (designated by the blue node). (D) A ctenophore. (E) A sponge.

@@

The published analyses considered here differ in a variety of ways, including the model of molecular evolution, gene selection, and ingroup and outgroup species selection.

The models used in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models, with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodate total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices. As a result, the genes selected for analysis vary widely (Fig. 2, Extended Data Fig. 1). To explore this variation in gene sampling, we annotated each gene against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table 1). These annotations show XXX. We identified the ribosomal protein and BUSCO genes [@waterhouse2017busco] in each data matrix (Extended Data Fig. 2). The animal BUSCO dataset contains 978 genes that tend to occur as a single copy used to assess the completeness of genome assemblies [@waterhouse2017busco]. All matrices have than 50% of the BUSCO genes, and several less than 20%. The fact that so many BUSCO genes are not in these matrices reflects that the gene selection criteria include many additional factors, such as high expression (since so many phylogenomic data are based on transcriptome data) and features of molecular evolution.

@@

**Fig. 2.** An overview of previous studies. (A) A time-series plot to show the main conclusions of previous 13 phylogenomic studies that tested Ctenophora-sister (designated by the orange node) and Porifera-sister (designated by the green node). (B) Matrix taxa and gene sampling for each study. Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

Ingroup taxon sampling across clades varies widely between studies (Fig. 2, Extended Data Fig. 1). Sampling of ingroup taxa (animals) was initially biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups. Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig. 1).

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).

@@

Our summary of previous phylogenetic analyses (Fig. 3A) show that Ctenophora-sister is supported over analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem (XX), but draws on a more systematic overview. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because CAT is so parameter rich, many analyses that use a model with CAT do not converge. Some of the uncovered analyses have an unresolved animal root. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across analyses. This extensive variation can make it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility than other maximum likelihood (ML) tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C10-C60 equilibrium frequencies [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites. There are some implementation differences. For example, CAT estimates the number of distinct site categories, the assignment of sites to those categories, and the equilibrium frequencies of each of those categories, while the C models specify the number of categories *a priori* (the C10 model as 10 categories, etc...).

@@

For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling of to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.

For each matrix, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] with BIC in IQ-TREE (except only LG+C60 models are used in 3 large matrices). In all cases, models with C60 fit these matrices better than the site-homogeneous models. We then inferred support under the best-fit model from IQ-TREE (Extended Data Table 2). We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified for Choanoflagellata taxon sampling (Extended Data Table 2).

All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the Ctenophora-sister hypothesis (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two deifferent means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses that lead to support for Porifera-sister.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to infer number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60. Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This requires is a very large number of additional parameters that must be estimated from the data. When ModelFinder is used to select the number of categories in IQ-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to BIC criteria (Supplementary Table 5). These findings are consistent with the concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of categories. When we sweep the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.

Published studies have found that the GTR-CAT model often has a better fit than Poisson-CAT, and investigators have suggested that it is more appropriate than the Poisson-CAT model for this question (XX citations). Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on matrices with Choanimalia sampling. We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model. Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.

These results further clarify when analyses support Porifera-sister. In our analyses, outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a Poisson rather than GTR exchange matrix must be used. Analyses under other conditions either support Ctenophora-sister or are unresolved.

@@

To further explore the variation described above, we quantify the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 3). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$GLS) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models (LG or WAG in both IQ-TREE or PhyloBayes). Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in site-homogeneous models. This finding is largely consistent with the the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

@@

Other approaches have been taken to addressing heterogeneity in equilibrium frequencies. For example, Feuda *et al.* [@Feuda:2017ew] employed recoding methods that recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. Our own analyses (Supplementary text, Supplementary Fig. XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

Resolving the placement of the root in the animal phylogeny has proven particularly difficult. By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some analyses. This demonstrates and important interaction between model selection and taxon sampling. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency. Site homogenous models have a single category that is applied to all sites. C models, as implemented in IQ-TREE, have 10-16 categories. CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT. When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories. As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig. 4). As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.

@@

Overcoming the challenges that face this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic perspective. In particular, there are very few chromosome-level genome assembles outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and greatly reduce the possibility of contamination in any of the gene sequence datasets. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

>>>>

There has been considerable recent debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in a many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 13 previous phylogenomic analyses and performing new standardized analyses. Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Our new analyses are consistent with this. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for phylogenetic methods development.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative. Since then, many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result. But the extensive technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity. This methodological variance is often confounded, for example when a particular model is only available in one software package.

To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 13 studies (Extended Data Table 1). These include all such phylogenomic studies of amino acid sequence data published before 2020 for which we could obtain data matrices with gene partition annotations.

![](figures/Figure_overview.png)
**Fig. 1.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta. (D) A ctenophore. (E) A sponge.

@@

The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models, with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodate total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices. As a result, the paritcular genes selected for analysis vary widely (Fig. 2, Extended Data Fig. 1). The composition of of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

**Fig. 2.** An overview of previous studies. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister ( orange nodes) or Porifera-sister (green nodes). (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

Ingroup taxon sampling also varies widely between studies (Fig. 2, Extended Data Fig. 1). Sampling of ingroup taxa (animals) was initially biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 1). This variation is in part because newer sequencing technologies in more recent papers are usually not applied to the exact same species that were in earlier papers.

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup subtree attaches to the ingroup subtree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [@graham2002rooting]. Most studies addressing the animal root have removed more distantly related outgroup taxa in some analyses to explore the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanimalia (animals plus most closely related Choanoflagellatea), Holozoa (Choanimalia plus more distantly related Holozoa) and Opisthokonta (Holozoa + Fungi).

@@

Our summary of previous phylogenetic analyses (Fig. 3A) show that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem (XX), but draws on a more systematic overview. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation can make it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C10-C60 equilibrium frequencies [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.

@@

For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_processed %>% filter(status=="new") %>% nrow()` matrices.

For all but the three largest matrices, we tested the relative fit of a variety of models, both with and without C60 accommodation of site heterogeneity in equilibrium frequencies, using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQ-TREE. In all cases, models with C60 fit these matrices better than the site-homogeneous models. We then inferred support under the best-fit model (Extended Data Table 2), except for the three largest matrices where we used LG+C60. We then analyzed each matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60 (Extended Data Table 2).

All IQ-TREE analyses, apart from unresolved analyses of Moroz2014_3d and all Nosenko2013 matrices, supported the Ctenophora-sister hypothesis (Fig. 3B). No IQ-TREE analyses supported Porifera-sister, even analyses that restrict outgroup sampling to Choanoflagellata and use models with site-heterogeneous equilibrium frequencies (Fig. 3B lower right; Supplementary Table 3), the conditions under which published PhyloBayes CAT analyses recover strong support for Porifera-sister (Fig. 3A). To further verify this difference in a controlled manner we reran PhyloBayes analyses with CAT, using both Poisson and GTR substitution matrices, for some matrices that had led to support for Porifera-sister in published analyses. Consistent with published results, some of these supported Porifera-sister.

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two deifferent means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to infer the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60. Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. When ModelFinder is used to select the number of categories in IQ-TREE C models, it often selects a number far less than 60, suggesting that 60 is already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories. When we swept the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.

Published studies have found that the GTR-CAT model often has a better fit than Poisson-CAT, and investigators have suggested that it is more appropriate than the Poisson-CAT model for this question (XX citations). Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on matrices with Choanimalia sampling. We found that for both Whelan2017 and Philippe matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model. Moreover, we also found all results strongly supported Ctenophora-sister with GTR-nCAT60 models.

These results further clarify when analyses support Porifera-sister. Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a Poisson rather than GTR exchange matrix must be used. Analyses under other conditions either support Ctenophora-sister or are unresolved.

@@

To further explore the variation described above, we quantify the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 3). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

@@

Other approaches have been taken to addressing heterogeneity in equilibrium frequencies. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. Our own analyses (Supplementary text, Supplementary Fig. XX-XX) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem. By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses. This demonstrates an important interaction between model selection and taxon sampling. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency. Site homogenous models have a single category that is applied to all sites. C models, as implemented in IQ-TREE, have 10-16 categories. CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT. When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories. As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig. 4). As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.

@@

Overcoming the challenges that face this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context. In particular, there are very few chromosome-level genome assembles outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
